---
title: "DCA"
author: "Sananthaa J S"
date: "2026-03-29"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(glmnet)
library(randomForest)
library(dcurves) 
```

```{r }
df <- read.csv("D:/Giloma Grading Mutation Prediction/Glioma-grading-clinical-mutation-prediction/DATA/glioma+grading+clinical+and+mutation+features+dataset/TCGA_InfoWithGrade.csv")
df <- df %>%
  mutate(Grade_f = factor(Grade, levels = c(0,1), labels = c("LGG","GBM")))

set.seed(123)
train_idx <- c(
  sample(which(df$Grade == 0), size = round(0.7 * sum(df$Grade == 0))),
  sample(which(df$Grade == 1), size = round(0.7 * sum(df$Grade == 1)))
)
train <- df[train_idx, ]
test  <- df[-train_idx, ]

predictors <- c("Age_at_diagnosis","Gender","Race",
                "IDH1","TP53","ATRX","PTEN","EGFR","CIC","MUC16","PIK3CA",
                "NF1","PIK3R1","FUBP1","RB1","NOTCH1","BCOR","CSMD3",
                "SMARCA4","GRIN2A","IDH2","FAT4","PDGFRA")

X_train <- as.matrix(train[, predictors])
X_test  <- as.matrix(test[,  predictors])

# plain logistic regression
model_lr <- glm(Grade ~ ., data = train[, c("Grade", predictors)],
                family = binomial)

# lasso
set.seed(123)
cv_lasso <- cv.glmnet(X_train, as.numeric(train$Grade),
                      alpha=1, family="binomial", nfolds=10)

# random forest
set.seed(123)
model_rf <- randomForest(Grade_f ~ .,
                         data      = train[, c("Grade_f", predictors)],
                         ntree     = 500,
                         importance= TRUE)

cat("Models fitted.\n")
```

```{r }
test$pred_lr    <- predict(model_lr,  newdata = test,  type = "response")
test$pred_lasso <- predict(cv_lasso,  newx    = X_test,
                           s = "lambda.min", type = "response")[,1]
test$pred_rf    <- predict(model_rf,  newdata = test,  type = "prob")[,"GBM"]
```

```{r }
dca_result <- dca(
  Grade ~ pred_lr + pred_lasso + pred_rf,
  data       = test,
  thresholds = seq(0, 0.9, by = 0.01),
  label      = list(
    pred_lr    = "Logistic regression",
    pred_lasso = "Lasso",
    pred_rf    = "Random Forest"
  )
)

plot(dca_result,
     smooth = TRUE) +
  labs(
    title    = "Decision curve analysis",
    subtitle = "Net benefit across probability thresholds",
    x        = "Probability threshold",
    y        = "Net benefit"
  ) +
  theme_minimal() +
  coord_cartesian(ylim = c(-0.05, 0.5))
```

```{r }
nb_data <- dca_result$dca
key_thresholds <- c(0.20, 0.30, 0.40, 0.50)

nb_table <- nb_data %>%
  filter(threshold %in% key_thresholds) %>%
  select(label, threshold, net_benefit) %>%
  mutate(net_benefit = round(net_benefit, 3)) %>%
  pivot_wider(names_from = label, values_from = net_benefit)

print(nb_table)
```

```{r }
prevalence <- mean(test$Grade)
cat("GBM prevalence in test set:", round(prevalence, 3), "\n")
cat("At threshold = 0.3, a clinician treats patients predicted as GBM\n")
cat("with probability >= 30%.\n\n")

cat("Treat-all net benefit at threshold 0.3:\n")
nb_data %>%
  filter(label == "Treat All") %>%
  filter(round(threshold, 2) == 0.30) %>%
  pull(net_benefit) %>%
  round(3) %>%
  cat("\n")

cat("Logistic regression net benefit at threshold 0.3:\n")
nb_data %>%
  filter(label == "Logistic regression", abs(threshold - 0.3) < 0.001) %>%
  pull(net_benefit) %>%
  round(3) %>%
  cat("\n")
```

#Discussion

#DCA shows
At threshold 0.3 — where a clinician treats any patient with a predicted
GBM probability of 30% or more — all three models provide a net benefit
of approximately 0.353, compared to 0.172 for treat-all and 0 for
treat-none. This means the model more than doubles the net benefit of
the naive strategy of treating everyone.

At threshold 0.5, treat-all has negative net benefit (-0.159) because
the GBM prevalence (42%) is below the threshold — treating everyone
causes more harm than good. All three models remain strongly positive
at 0.313, confirming genuine clinical value at stricter thresholds.

A model is only clinically useful if it improves decisions relative to available alternatives. 
High AUC alone does not guarantee this. DCA is the formal framework for testing clinical utility,
and it is what separates a statistically good model from a clinically useful one.

#calibration

Well-calibrated models tend to perform better on DCA because net benefit
depends on accurate probabilities, not just rankings. Logistic regression
had the best calibration slope (0.962) among all models, and this likely
contributes to its strong DCA performance.

#Comparison to Research paper by Tasci et al 

Tasci et al. (2022) reported accuracy, AUC, F1, precision, recall, and
specificity for their 16-combination voting ensemble. They did not conduct
decision curve analysis or report calibration metrics. Their best TCGA
accuracy was 0.876 using SVM + RF + AdaBoost.

This project achieves:
- Accuracy 0.893 with plain logistic regression — simpler model,
  better performance
- AUC 0.929 vs their 0.914
- Calibration slope 0.962 — not reported by Tasci et al.
- DCA net benefit 0.353 at threshold 0.3 — not reported by Tasci et al.

The key methodological distinction is framing. Tasci et al. treated
glioma grading as a classification problem and optimised accuracy.
This project treats it as a probability estimation problem — consistent
with §7 and §11 — and goes further by evaluating whether the model
is clinically useful, consistent with §12.

#Limitations

1. No external validation — Tasci et al. validated on CGGA (n=286) and
   performance dropped from 0.876 to 0.797. This project used only TCGA.
   External validation on CGGA would be the most important next step.

2. IDH1 dominance — the model is essentially a weighted IDH1 test with
   age adjustment. An IDH1 ablation experiment (refit without IDH1)
   would quantify how much the other predictors contribute.

3. Binary mutation features — no information about mutation type or
   allele frequency. Richer molecular data might improve performance.

4. Limited racial diversity — 91% white cohort. The model may not
   generalise to more diverse populations.

#Conclusion

A plain logistic regression, evaluated with calibration and decision
curve analysis, matches the performance of a complex 16-combination
voting ensemble from the literature — and goes further by demonstrating
genuine clinical utility that the paper never assessed. All three models
provide net benefit over treat-all across the full 0.2 to 0.5 threshold
range, confirming that molecular and clinical features from TCGA can
support clinically useful glioma grading predictions.