---
title: "Model Evaluation"
author: "Sananthaa J S"
date: "2026-03-29"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(glmnet)
library(randomForest)
library(pROC)
library(rms)
```


```{r }
df <- read.csv("D:/Giloma Grading Mutation Prediction/Glioma-grading-clinical-mutation-prediction/DATA/glioma+grading+clinical+and+mutation+features+dataset/TCGA_InfoWithGrade.csv")
df <- df %>%
  mutate(Grade_f = factor(Grade, levels = c(0,1), labels = c("LGG","GBM")))

set.seed(123)
train_idx <- c(
  sample(which(df$Grade == 0), size = round(0.7 * sum(df$Grade == 0))),
  sample(which(df$Grade == 1), size = round(0.7 * sum(df$Grade == 1)))
)
train <- df[train_idx, ]
test  <- df[-train_idx, ]

predictors <- c("Age_at_diagnosis","Gender","Race",
                "IDH1","TP53","ATRX","PTEN","EGFR","CIC","MUC16","PIK3CA",
                "NF1","PIK3R1","FUBP1","RB1","NOTCH1","BCOR","CSMD3",
                "SMARCA4","GRIN2A","IDH2","FAT4","PDGFRA")

X_train <- as.matrix(train[, predictors])
X_test  <- as.matrix(test[,  predictors])
y_train <- train$Grade
y_test  <- test$Grade

# plain logistic regression
model_lr  <- glm(Grade ~ ., data = train[, c("Grade", predictors)],
                 family = binomial)

# lasso
set.seed(123)
cv_lasso  <- cv.glmnet(X_train, y_train, alpha=1,
                       family="binomial", nfolds=10)

# ridge
set.seed(123)
cv_ridge  <- cv.glmnet(X_train, y_train, alpha=0,
                       family="binomial", nfolds=10)

# random forest
set.seed(123)
model_rf  <- randomForest(Grade_f ~ ., 
                          data      = train[, c("Grade_f", predictors)],
                          ntree     = 500,
                          importance= TRUE)

cat("All models fitted.\n")
```

```{r }
pred_lr    <- predict(model_lr,  newdata = test,   type = "response")
pred_lasso <- predict(cv_lasso,  newx    = X_test,
                      s = "lambda.min", type = "response")[,1]
pred_ridge <- predict(cv_ridge,  newx    = X_test,
                      s = "lambda.min", type = "response")[,1]
pred_rf    <- predict(model_rf,  newdata = test,   type = "prob")[,"GBM"]
```


```{r }
# training predictions
pred_lr_train    <- predict(model_lr, newdata = train, type = "response")
pred_lasso_train <- predict(cv_lasso, newx = X_train,
                            s = "lambda.min", type = "response")[,1]
pred_ridge_train <- predict(cv_ridge, newx = X_train,
                            s = "lambda.min", type = "response")[,1]
pred_rf_train    <- predict(model_rf, newdata = train, type = "prob")[,"GBM"]

auc_lr_train    <- auc(roc(y_train, pred_lr_train))
auc_lasso_train <- auc(roc(y_train, pred_lasso_train))
auc_ridge_train <- auc(roc(y_train, pred_ridge_train))
auc_rf_train    <- auc(roc(y_train, as.numeric(pred_rf_train)))

auc_lr_test     <- auc(roc(y_test, pred_lr))
auc_lasso_test  <- auc(roc(y_test, pred_lasso))
auc_ridge_test  <- auc(roc(y_test, pred_ridge))
auc_rf_test     <- auc(roc(y_test, pred_rf))

overfit_table <- tibble(
  Model        = c("Logistic regression","Lasso","Ridge","Random Forest"),
  Train_AUC    = round(c(auc_lr_train, auc_lasso_train,
                         auc_ridge_train, auc_rf_train), 3),
  Test_AUC     = round(c(auc_lr_test, auc_lasso_test,
                         auc_ridge_test, auc_rf_test),  3),
  Gap          = round(Train_AUC - Test_AUC, 3)
)

print(overfit_table)
```

```{r }
get_metrics <- function(prob, actual, threshold) {
  pred  <- ifelse(prob >= threshold, 1, 0)
  TP    <- sum(pred == 1 & actual == 1)
  TN    <- sum(pred == 0 & actual == 0)
  FP    <- sum(pred == 1 & actual == 0)
  FN    <- sum(pred == 0 & actual == 1)
  list(
    sensitivity = TP / (TP + FN),
    specificity = TN / (TN + FP),
    accuracy    = (TP + TN) / length(actual)
  )
}
```


```{r }
models <- list(
  "Logistic regression" = pred_lr,
  "Lasso"               = pred_lasso,
  "Ridge"               = pred_ridge,
  "Random Forest"       = pred_rf
)

# metrics at both thresholds
results_05 <- map_dfr(names(models), function(m) {
  mt <- get_metrics(models[[m]], y_test, 0.5)
  tibble(Model = m, Threshold = 0.5,
         Sensitivity = round(mt$sensitivity, 3),
         Specificity = round(mt$specificity, 3),
         Accuracy    = round(mt$accuracy,    3))
})

results_03 <- map_dfr(names(models), function(m) {
  mt <- get_metrics(models[[m]], y_test, 0.3)
  tibble(Model = m, Threshold = 0.3,
         Sensitivity = round(mt$sensitivity, 3),
         Specificity = round(mt$specificity, 3),
         Accuracy    = round(mt$accuracy,    3))
})

cat("=== Threshold = 0.5 (default) ===\n")
print(results_05)
cat("\n=== Threshold = 0.3 (clinical — prioritise sensitivity) ===\n")
print(results_03)
```

```{r brier}
brier <- tibble(
  Model       = names(models),
  Brier_score = round(map_dbl(models, ~ mean((.x - y_test)^2)), 4)
)
print(brier)
```

```{r }
get_cal <- function(prob, actual) {
  fit <- val.prob(prob, actual, pl = FALSE)
  list(slope = round(fit["Slope"], 3),
       intercept = round(fit["Intercept"], 3))
}

cal_lr    <- get_cal(pred_lr,    y_test)
cal_lasso <- get_cal(pred_lasso, y_test)
cal_ridge <- get_cal(pred_ridge, y_test)
cal_rf    <- get_cal(pred_rf,    y_test)

cal_table <- tibble(
  Model     = names(models),
  Cal_slope = c(cal_lr$slope, cal_lasso$slope,
                cal_ridge$slope, cal_rf$slope),
  Cal_intercept = c(cal_lr$intercept, cal_lasso$intercept,
                    cal_ridge$intercept, cal_rf$intercept)
)
print(cal_table)
```

---

## 8. Final comparison table — §11, §13

```{r final_table}
final_table <- tibble(
  Model       = names(models),
  AUC         = round(c(auc_lr_test, auc_lasso_test,
                        auc_ridge_test, auc_rf_test), 3),
  Brier       = brier$Brier_score,
  Cal_slope   = cal_table$Cal_slope,
  Cal_int     = cal_table$Cal_intercept,
  Sens_05     = results_05$Sensitivity,
  Spec_05     = results_05$Specificity,
  Sens_03     = results_03$Sensitivity,
  Spec_03     = results_03$Specificity
)

print(final_table)
```

---

## 9. Overlaid ROC curves — all four models — §7

```{r roc_overlay}
roc_lr    <- roc(y_test, pred_lr)
roc_lasso <- roc(y_test, pred_lasso)
roc_ridge <- roc(y_test, pred_ridge)
roc_rf    <- roc(y_test, pred_rf)

plot(roc_lr,
     col  = "#3B8BD4", lwd = 2,
     main = "ROC curves — all models")
lines(roc_lasso, col = "#E24B4A",  lwd = 2)
lines(roc_ridge, col = "#EF9F27",  lwd = 2)
lines(roc_rf,    col = "#1D9E75",  lwd = 2)
abline(a = 0, b = 1, lty = 2, col = "grey60")

legend("bottomright",
       legend = c(
         paste0("Logistic regression (", round(auc_lr_test,  3), ")"),
         paste0("Lasso (",               round(auc_lasso_test,3), ")"),
         paste0("Ridge (",               round(auc_ridge_test,3), ")"),
         paste0("Random Forest (",       round(auc_rf_test,  3), ")")
       ),
       col = c("#3B8BD4","#E24B4A","#EF9F27","#1D9E75"),
       lwd = 2, cex = 0.85)
```

```{r best_calibration}
val.prob(pred_lr, y_test)
title("Calibration — best model (logistic regression)", line = 0.5)
```

NOTES:

The overfitting table shows whether any model memorised training data
The 0.3 threshold comparison shows sensitivity gains at cost of specificity
clinically motivated because missing GBM is worse than over-flagging
Plain logistic regression wins on calibration — best for clinical use
All four models have very similar AUC — the problem is learnable with simple methods, complexity does not help here.